Psychological Needs During Intergroup Contact

### 2021-10-25 manuscript.tex
notes added
Updates from Overleaf

% Relevance
Conflict between social groups and their individual members remains a prevalent feature of the modern human condition. Prejudices, discrimination, and animosities continue to plague societies around the world. One of the main ameliorations proposed by social psychologists has been positive intergroup contact. Based on Allport's (\citeyear{Allport1954b}) contact hypothesis, a plethora of studies and interventions have worked on bringing different social groups together\footnote{Cite interventions of British ministries.}.

\section{The Present Research}

@@

>>>>

% Relevance
Conflict between social groups and their individual members remains a prevalent feature of the modern human condition. Prejudices, discrimination, and animosities continue to plague societies around the world. One of the main ameliorations proposed by social psychologists has been positive intergroup contact. Based on Allport's \citeyear{Allport1954b} contact hypothesis, a plethora of studies and interventions have worked on bringing different social groups together\footnote{Cite interventions of British ministries.}. 
% Yet, despite ...

In its most essential interpretation, the intergroup contact hypothesis postulates that frequent and positive contact with an out-group reduces prejudice and increases favorable attitudes towards the other group \citep[e.g.,][]{Hewstone1996, Pettigrew1998}. A key condition for these contact benefits, has been that the interaction is indeed perceived as positive --- making the interaction quality a crucial mechanism of inter-group contact \citep[e.g.,][]{MacInnis2015}. It is widely accepted that equal status, common goals, collaboration, and structural support during the interaction form the optimal conditions for a positive contact \citep[Allport's Optimal Contact conditions,][]{Allport1954b}. And indeed a major meta-analytic review showed, that intergroup contact benefits were larger when Allport's conditions were met. However, the meta-analysis also showed that contact resulted in more positive intergroup relations even when Allport's conditions were not met. It, thus, remains unclear why and how exactly Allport's conditions work.

In this article, we propose that Allport's optimal contact conditions are effective in creating positive contact because they constitute common psychological needs of the interacting individuals. If this is indeed the case, the impact of fulfilling other fundamental needs should produce similar beneficial contact results. And more importantly, the satisfaction of key situation needs during the interaction should predict the positive contact effects most strongly.

\faKeyboard\ We should probably add a short section on past literature that has examined psychological needs in intergroup contact in general \citep[e.g.,][]{Shnabel2008} as well as for particular groups, including migrants \citep[e.g.,][]{Celebi2017}.

\noindent We should probably also include:
\begin{displayquote}
    Psychological needs might “explain the asymmetry of experiences and outcomes in intergroup interactions and identify an essential element for interventions: To achieve truly constructive intergroup relations, it is important that intergroup exchanges meet the psychological needs of both majority- and minority-group members” \\
    \hfill --- \citep[][p. 611]{Dovidio2017}
\end{displayquote}

We probably need to note that there are really good papers on the likely mechanisms of intergroup contact \citep[e.g.,][]{Brown2005}. But they don't discuss Allport's conditions.

Plus, recent work Nurit Shnabel.

\section{The Present Research}
\faKeyboard\ To be written out.

@@

Based on this we formulated four main hypotheses:

\begin{hyp}[H\ref{hyp:contact}] \label{hyp:contact}
Based on the most general understanding of the contact hypothesis, an increase in frequency and quality of contact should jointly account for changes in more favorable outgroup attitudes.
\end{hyp}

\begin{subhyp}[H\ref{hyp:contactFreq}] \label{hyp:contactFreq}
\addtolength{\leftskip}{2.5em}
Participants with more intergroup interactions should have a more favorable outgroup attitudes.
\end{subhyp}

\begin{subhyp}[H\ref{hyp:contactDummy}] \label{hyp:contactDummy}
\addtolength{\leftskip}{2.5em}
Outgroup attitudes should be more positive after an intergroup interaction compared to a non-outgroup interaction.
\end{subhyp}

\begin{subhyp}[H\ref{hyp:contactFreqQual}] \label{hyp:contactFreqQual}
\addtolength{\leftskip}{2.5em}
Participants with more intergroup interactions should have a more favorable outgroup attitudes depending on the average interaction quality.
\end{subhyp}

\begin{hyp}[H\ref{hyp:AllportsConditions}] \label{hyp:AllportsConditions}
Based on Allport's optimal contact conditions, intergroup interactions with equal status, common goals, collaboration, and structural support should predict more favorable outgroup attitudes due to more positive interaction quality perceptions.
\end{hyp}

\setcounter{subhyp}{0}
\begin{subhyp}[H\ref{hyp:AllportsPred}] \label{hyp:AllportsPred}
\addtolength{\leftskip}{2.5em}
Based on Allport's optimal contact conditions, outgroup attitudes should be more favorable after intergroup interactions with equal status, common goals, collaboration, and structural support.
\end{subhyp}

\begin{subhyp}[H\ref{hyp:AllportsQuality}] \label{hyp:AllportsQuality}
\addtolength{\leftskip}{2.5em}
Based on past research on the role of interaction quality, interaction quality should be more perceived as more favorable after intergroup interactions with equal status, common goals, collaboration, and structural support.
\end{subhyp}

\begin{subhyp}[H\ref{hyp:AllportsQualityMediation}] \label{hyp:AllportsQualityMediation}
\addtolength{\leftskip}{2.5em}
Based on past research on the role of interaction quality, the variance explained in outgroup attitudes by Allport's optimal contact should to a large extend be assumed by interaction quality.
\end{subhyp}

\begin{hyp}[H\ref{hyp:keyNeed}] \label{hyp:keyNeed}
Based on our proposal, intergroup interactions with higher situational core need fulfillment should predict more favorable outgroup attitudes due to more positive interaction quality perceptions.
\end{hyp}

\setcounter{subhyp}{0}
\begin{subhyp}[H\ref{hyp:keyNeedPred}] \label{hyp:keyNeedPred}
\addtolength{\leftskip}{2.5em}
Outgroup attitudes should be more favorable after intergroup interactions with high key need fulfillment.
\end{subhyp}

\begin{subhyp}[H\ref{hyp:keyNeedQual}] \label{hyp:keyNeedQual}
\addtolength{\leftskip}{2.5em}
Interaction Quality should be perceived as more positive after intergroup interactions with higher key need fulfillment.
\end{subhyp}

\begin{subhyp}[H\ref{hyp:keyNeedMediation}] \label{hyp:keyNeedMediation}
\addtolength{\leftskip}{2.5em}
The variance explained in outgroup attitudes by key need fulfillment should to a large extend be assumed by interaction quality.
\end{subhyp}

\begin{subhyp}[H\ref{hyp:keyNeedContactType}] \label{hyp:keyNeedContactType}
\addtolength{\leftskip}{2.5em}
The effect of key need fulfillment on outgroup attitudes should be specific to intergroup interactions and not be due to need fulfillment in general. Thus, the effect of key need fulfillment on outgroup attitudes should stronger for intergroup interact than for ingroup interactions. 
\end{subhyp}

\begin{subhyp}[H\ref{hyp:keyNeedSDT}] \label{hyp:keyNeedSDT}
\addtolength{\leftskip}{2.5em}
The effect of key need fulfillment on outgroup attitudes should be persist even when taking other fundamental psychological needs into account. Thus, the effect of key need fulfillment on outgroup attitudes should remain strong even after controlling for autonomy, competence, and relatedness fulfillment during the interaction (cf., self-determination theory). 
\end{subhyp}

\begin{hyp}[H\ref{hyp:comparison}] \label{hyp:comparison}
Based on our proposal, intergroup interactions with higher situational core need fulfillment should predict outgroup attitudes at least as well as Allport's conditions.
\end{hyp}

\setcounter{subhyp}{0}
\begin{subhyp}[H\ref{hyp:compModel}] \label{hyp:compModel}
\addtolength{\leftskip}{2.5em}
The need model (H\ref{hyp:keyNeedPred}) should predict more variance in outgroup attitudes than the model based on Allport's conditions (H\ref{hyp:AllportsPred}).
\end{subhyp}

\begin{subhyp}[H\ref{hyp:compTogether}] \label{hyp:compTogether}
\addtolength{\leftskip}{2.5em}
The  effect of key need fulfillment on outgroup attitudes should  persist even when taking other Allport's conditions into account. Thus, the effect of key need fulfillment on outgroup attitudes should remain strong even after controlling for equal status, common goals, collaboration, and structural support.
\end{subhyp}

\faKeyboard\ all assumptions were fine (see Online Supplementary Materials B).
